---
title: "assignment 3"
author: "Bill Co"
date: "2025-03-10"
output: html_document
---

#Load the libraries
```{r}
library(MTPS)

data(HIV)

library(tidyverse)

library(caret)

library(tree)

library(rpart)
```

# Settings
```{r}
set.seed(0)

n.sim = 10
```

```{r}
XX <- as.data.frame(XX)

YY <- as.data.frame(YY)

drug.names <- colnames(YY)
```

## Define binary outcomes based on Cutoffs (provided by the assignment)
```{r}
yBin <- as.matrix(YY)
cutoffs <- c(2,3,3,1.5,1.5) # cutoff value to be used to
#define drug resistance
for(ii in 1:5) yBin[,ii] <- (10^yBin[,ii] < cutoffs[ii])*1
```

#Because resampling with large N can maintain the ratio between classes, so I am going to re-generate this index matrix only once
#Matrix is (10 x 5) x 1245/5. 

#-> There are 1246 obs but I sacrificed one for an easier matrix creation, so 1245 observations.
#-> Each of the 10 simulations will have 5 rows for 5 fold CV.
```{r}
indices <- matrix(sample(seq(1, 1246), n.sim * 1245, replace = T), nrow = n.sim * 5, ncol = 1245/5)
dim(indices)
```


```{r}
confusion.metrics <- function(predictions, actual) {
  
  cm <- table(actual, predictions)
  TN <- cm[1,1]
  FP <- cm[1,2]
  FN <- cm[2,1]
  TP <- cm[2,2]
  
  accuracy <- (TN  + TP) / (TN + FN + FP + TP)
  mcr <- 1 - accuracy
  precision <- TP / (TP + FP)
  recall <- TP / (TP + FN)
  f1 <- (2*TP) / (2*TP + FP + FN)
  
  c(accuracy, mcr, precision, recall, f1)
}
```




```{r}
logistic.performances <- matrix(0, nrow=n.sim, ncol=5 * 5)
lda.performances      <- matrix(0, nrow=n.sim, ncol=5 * 5)
elastic.performances  <- matrix(0, nrow=n.sim, ncol=5 * 5)
tree.performances     <- matrix(0, nrow=n.sim, ncol=5 * 5)
forest.performances   <- matrix(0, nrow=n.sim, ncol=5 * 5)
  
#performance .id is the index of column, the way a performance matrix is designed is
#n rows = n simulations
#5 groups of cols for 5 drugs, each has 5 cols for 5 metrics (accuracy is not used due to imbalance class), so n.cols in total = 5*5

performance.id <- 1
  
for (drug.id in 1:5) {
  
  metric.id <- 1
  
  #because there are 5 drugs
  current_outputs <- yBin[,drug.id]
  
  #For logistic model, I'll use Lasso (L1) regularization, thus I will find out the best overall lambda before starting simulations
  
  cv.logistic <- cv.glmnet(as.matrix(XX), current_outputs, family = 'binomial', alpha = 1)
  cv.elastic <- cv.glmnet(as.matrix(XX), current_outputs, family = 'binomial', alpha = 0.5)
  
  print(paste('Assessing drug: ', drug.names[drug.id]))
  
  for (sim in seq(1, 5 * n.sim, 5)) {
    
    #because there are 10 simulations,
    #each with 5 rows(folds), so I extract the sub-matrix for simplicity.
    
    cur.sim.indices <- indices[sim: (sim+4), ] 
    
    
    #This metric matrix will be refreshed after every simulation
    
    metric.base <- matrix(0, nrow = 5, ncol = 5)
    
    lda.success <- 5
    logistic.success <- 5
    elastic.success <- 5
    tree.success <- 5
    forest.success <- 5
    
    
    for (test.id in 1:5) {
    
      #because there are 5 folds in a Cross Validation
  
      X.train <- as.matrix( XX[cur.sim.indices[-test.id, ], ] )
      X.test  <- as.matrix( XX[cur.sim.indices[ test.id, ], ] )
      
      y.train <- current_outputs[cur.sim.indices[-test.id, ]]
      y.test  <- current_outputs[cur.sim.indices[test.id, ]]
      
      
      ### Put model fittings here
      
      #1. Logistic with L1
      fitted.logistic <- glmnet(X.train, y.train, family = 'binomial', alpha = 1, lambda = cv.logistic$lambda.min)
      
      if (fitted.logistic$lambda != Inf) {
      
        logistic.probs <- predict(fitted.logistic, newx = X.test, type = 'response')
        
        logistic.metrics <- confusion.metrics(logistic.probs > .5, y.test)
        
        metric.base[1, ] <- metric.base[1, ] + logistic.metrics
      }
      
      else {
        
        logistic.success <- logistic.success - 1
      }
      
      #2. LDA
      
      tryCatch({
          # Fit LDA model
          fitted.lda <- lda(X.train, y.train)
          
          lda.predictions <- predict(fitted.lda, newdata = X.test)
      
          lda.metrics <- confusion.metrics(lda.predictions$class, y.test)
          
          metric.base[2, ] <- metric.base[2, ] + lda.metrics
          
      }, error = function(e) {
          lda.success <- lda.success - 1
      })
     
      
      #3. Elastic Net
      fitted.elastic <- glmnet(X.train, y.train, family = 'binomial', alpha = 0.5, lambda = cv.elastic$lambda.min)
        
        if (fitted.elastic$lambda != Inf) {
        
          elastic.probs <- predict(fitted.elastic, newx = X.test, type = 'response')
          
          elastic.metrics <- confusion.metrics(elastic.probs > .5, y.test)
          
          metric.base[3, ] <- metric.base[3, ] + elastic.metrics
          
        }
        
        else {
          
          elastic.success <- elastic.success - 1
        }
    
      #Dataframe for tree based
      XXX <- cbind(y.train, as.data.frame(X.train))
      
      #4. Trees
      
      fitted.tree <- tree(y.train ~ ., data=XXX)
      
      prune <- cv.tree(fitted.tree)
  
      best.size <- prune$size[which.min(prune$dev)]
      
      fitted.tree <- prune.tree(fitted.tree, best = best.size)
      
      y <- predict(fitted.tree, newdata = as.data.frame(X.test))
      
      tree.metrics <- confusion.metrics(y > .5, y.test)
      
      metric.base[4, ] <- metric.base[4, ] + tree.metrics

    
      
      
      #5. Random Forest
      
      forest <- rpart(y.train ~ ., data = XXX, method = "class")
      
      predictions <- predict(forest, as.data.frame(X.test))
      
      forest.metrics <- confusion.metrics(predictions[,2] > .5, y.test)
      
      metric.base[5, ] <- metric.base[5, ] + forest.metrics

    }
    
    
    ### Model performance
    # Save to matrix
    ###
    metric.base[1,] <- metric.base[1,] / logistic.success
    metric.base[2,] <- metric.base[2,] / lda.success #Divided by successful n folds
    metric.base[3,] <- metric.base[3,] / elastic.success #Divided by successful n folds
    metric.base[4,] <- metric.base[4,] / tree.success
    metric.base[5,] <- metric.base[5,] / forest.success
  
  
    
    # Save to simulation metric matrices
    
    logistic.performances[metric.id, performance.id : (performance.id + 4)] <- metric.base[1, ]
    lda.performances[metric.id, performance.id : (performance.id + 4)] <- metric.base[2, ]
    elastic.performances[metric.id, performance.id : (performance.id + 4)] <- metric.base[3, ]
    tree.performances[metric.id, performance.id : (performance.id + 4)] <- metric.base[4, ]
    forest.performances[metric.id, performance.id : (performance.id + 4)] <- metric.base[5, ]
    
    metric.id <- metric.id + 1
  }
  
  performance.id <- performance.id + 5 #Each performance_id is for each drug
  
}
```


#We have to write a another loop for the MTPS because it requires more than 1 class at fitting.
```{r}
mtps.performances  <-  matrix(0, nrow=n.sim, ncol=5 * 5)
metric.id <- 1

for (sim in seq(1, 5 * n.sim, 5)) {
  
    cur.sim.indices <- indices[sim: (sim+4), ] 
    
    #This metric matrix will be refreshed after every simulation
    
    for (test.id in 1:5) {
      
      performance.id <- 1
    
      #because there are 5 folds
  
      X.train <- as.matrix( XX[cur.sim.indices[-test.id, ], ] )
      X.test  <- as.matrix( XX[cur.sim.indices[ test.id, ], ] )
      
      y.train <- as.matrix( yBin[cur.sim.indices[-test.id, ], ] )
      y.test  <- as.matrix( yBin[cur.sim.indices[ test.id, ], ] )
      
      #Fit the mtps model
      fit.rs <- MTPS(xmat = X.train, ymat= y.train, family='binomial', cv=FALSE, residual = TRUE, method.step1 = rpart1, method.step2 = lm1)
      
      #Prediction for 5 drugs
      probs <- predict(fit.rs, X.test, type='response')
      
        for (drug in 1:5) { 
        
          #Record performances
          mtps.metrics <- confusion.metrics(probs[,drug] > .5, y.test[,drug])
          
          mtps.performances[metric.id, performance.id: (performance.id + 4)] <- mtps.metrics
          
          performance.id <- performance.id + 5
        }
      
    }
  
    metric.id <- metric.id + 1
}
```


#Visualization

## Because this data is very class-imbalanced, accuracy can be misleading so I am not including it


```{r}
all.metric <- function(metric, drug) {
  
  m <- matrix(c(logistic.performances[,metric + (drug-1) * 5], 
           lda.performances[,metric + (drug-1) * 5],
           elastic.performances[,metric + (drug-1) * 5],
           tree.performances[,metric + (drug-1) * 5],
           mtps.performances[,metric + (drug-1) * 5],
           forest.performances[,metric + (drug-1) * 5]),
           ncol=6)
  
  colnames(m) <- c("Logist.", "LDA", "Elastic", "Tree", "MTPS", "Forest")
  
  m
}
```

```{r}
metric.names <- c("Accuracy", "Mcr", "Precision", "Recall", "F1-Score")
```


#1. Box plot of metrics for each drug

## a. Drug ABC

```{r fig.height=5, fig.width=15}
cur.drug = 1

par(mfrow = c(1, 3),cex.axis = 1, cex.lab = 1, las = 1)

for (i in 3:5) { 
  x <- all.metric(i, cur.drug)

  
  boxplot(x, ylab = metric.names[i],
            main = paste(metric.names[i]," for drug", drug.names[cur.drug]),
            col = c("lightblue", "lightgreen", "lightcoral", "lightgoldenrodyellow", "orange", "grey"))
}
```
## b. Drug 3TC
```{r fig.height=5, fig.width=15}
cur.drug = 2

par(mfrow = c(1, 3),cex.axis = 1, cex.lab = 1, las = 1)

for (i in 3:5) { 
  x <- all.metric(i, cur.drug)

  
  boxplot(x, ylab = metric.names[i],
            main = paste(metric.names[i]," for drug", drug.names[cur.drug]),
            col = c("lightblue", "lightgreen", "lightcoral", "lightgoldenrodyellow", "orange", "grey"))
}
```

## c. Drug AZT

```{r fig.height=5, fig.width=15}
cur.drug = 3

par(mfrow = c(1, 3),cex.axis = 1, cex.lab = 1, las = 1)

for (i in 3:5) { 
  x <- all.metric(i, cur.drug)

  
  boxplot(x, ylab = metric.names[i],
            main = paste(metric.names[i]," for drug", drug.names[cur.drug]),
            col = c("lightblue", "lightgreen", "lightcoral", "lightgoldenrodyellow", "orange", "grey"))
}
```

# Drug D4T

```{r fig.height=5, fig.width=15}
cur.drug = 4

par(mfrow = c(1, 3),cex.axis = 1, cex.lab = 1, las = 1)

for (i in 3:5) { 
  x <- all.metric(i, cur.drug)

  
  boxplot(x, ylab = metric.names[i],
            main = paste(metric.names[i]," for drug", drug.names[cur.drug]),
            col = c("lightblue", "lightgreen", "lightcoral", "lightgoldenrodyellow", "orange", "grey"))
}
```

# Drug DDI
```{r fig.height=5, fig.width=15}
cur.drug = 5

par(mfrow = c(1, 3),cex.axis = 1, cex.lab = 1, las = 1)

for (i in 3:5) { 
  x <- all.metric(i, cur.drug)

  
  boxplot(x, ylab = metric.names[i],
            main = paste(metric.names[i]," for drug", drug.names[cur.drug]),
            col = c("lightblue", "lightgreen", "lightcoral", "lightgoldenrodyellow", "orange", "grey"))
}
```

#2. Box plot of metrics for all drugs combined

```{r}
group.logistic.performance <- sapply(1:5, function(i) rowSums(logistic.performances[, seq(i, 25, by = 5)])) / 5

group.lda.performance <- sapply(1:5, function(i) rowSums(lda.performances[, seq(i, 25, by = 5)])) / 5

group.elastic.performance <- sapply(1:5, function(i) rowSums(elastic.performances[, seq(i, 25, by = 5)])) / 5

group.tree.performance <- sapply(1:5, function(i) rowSums(tree.performances[, seq(i, 25, by = 5)])) / 5

group.forest.performance <- sapply(1:5, function(i) rowSums(forest.performances[, seq(i, 25, by = 5)])) / 5

group.mtps.performance <- sapply(1:5, function(i) rowSums(mtps.performances[, seq(i, 25, by = 5)])) / 5
```

```{r}
group.metric <- function(metric) {
  
  m <- matrix(c(group.logistic.performance[,metric], 
             group.lda.performance[,metric],
             group.elastic.performance[,metric],
             group.tree.performance[,metric],
             group.mtps.performance[,metric],
             group.forest.performance[,metric]), 
             ncol=6)
    
  colnames(m) <- c("Logist.", "LDA", "Elastic", "Tree", "MTPS", "Forest")
  
  m
}
```

```{r fig.height=5, fig.width=15}
par(mfrow = c(1, 3),cex.axis = 1, cex.lab = 1, las = 1)

for (i in 3:5) {
  x <- group.metric(i)


  boxplot(x, ylab = metric.names[i],
            main = paste("Combined ", metric.names[i]),
            col = c("lightblue", "lightgreen", "lightcoral", "lightgoldenrodyellow", "orange", "grey"))
}
```

# Hypothesis Testing

We will conduct the Wilcoxon paired test for Precision between Logistic and Elastic Nets
```{r}
median(group.logistic.performance[, 3])
median(group.elastic.performance[, 3])
wilcox.test(group.logistic.performance[, 3], group.elastic.performance[, 3], paired = T)
```

We will conduct the Wilcoxon paired test for Recall between Logistic and Elastic Nets

```{r}
median(group.logistic.performance[, 4])
median(group.elastic.performance[, 4])
wilcox.test(group.logistic.performance[, 4], group.elastic.performance[, 4], paired = T)
```

We will conduct the Wilcoxon paired test for F1-Score between Logistic and Elastic Nets

```{r}
median(group.logistic.performance[, 5])
median(group.elastic.performance[, 5])
wilcox.test(group.logistic.performance[, 5], group.elastic.performance[, 5], paired = T)
```

